---
title: SA
author: Doug
layout: default
---


```{r, include=FALSE}
library(tidyverse)
base_name = knitr::current_input() %>% str_split('\\.', simpl=TRUE) %>% `[`(1)
knitr::opts_chunk$set(
  cache.path=str_c('_knitr_cache/', base_name, '/'),
  fig.path=str_c('figure/', base_name, '/'),
  dpi=300,
  fig.width=4,
  fig.height=3,
  fig.cap=''
)
```

**TODO:**

- Why the reordering with `unique_fingerprint`?
- find better solution to make the output file available
- methodological details, reference and comment on results of FAST99
- fix negative values for `I` produced by emulator (emulate log(I)?)
- do everything with `mo_gp`



**Disclaimer:** This code is an example only, and not (yet) a serious analysis.
Results of the sensitivity analysis will change - perhaps dramatically - when
sensible ranges for the parameters are used.


This code takes a small ensemble of runs of MetaWards runs and fits a Gaussian
Process emulator to the maximum number of infections in each run. The code then
does a sensitivty analysis using the FASTT99 algorithm, and emulated output.
Finally, it looks at the one-at-a-time sensitivity using emulated output.


## Load Packages

```{r}
library(tidyverse)
library(sensitivity)
library(DiceKriging)
source("https://raw.githubusercontent.com/dougmcneall/packages-git/master/emtools.R")
```

## Read the design matrix 

Load the design file created in [metawards_design.md](metawards_design):

```{r}
# Need to fix the parameter names
design_file = 'https://raw.githubusercontent.com/dougmcneall/covid/master/experiments/2020-05-07-sensitivity-analysis/design.csv'
X <- read.csv(design_file, sep = "")
parnames = colnames(X)
```

## Read and summarise the model runs

Load the output file created in [metawards_design.md](metawards_design). The output file used here can be downloaded from <https://github.com/dougmcneall/covid/blob/master/experiments/2020-05-07-sensitivity-analysis/output/results.csv.bz2>

```{r, eval=TRUE, echo=FALSE}
# A container for all the data 
# Each row has a "fingerprint" that contains the
# values of all the changed parameters, and the values of the parameters are also
# given.  This alters the order of the parameters.
dat <- read.csv('~/Downloads/results.csv.bz2')
```
```{r, eval=FALSE, echo=TRUE}
# A container for all the data 
# Each row has a "fingerprint" that contains the
# values of all the changed parameters, and the values of the parameters are also
# given.  This alters the order of the parameters.
dat <- read.csv('results.csv.bz2')
```

```{r}
unique_fingerprint = unique(dat$fingerprint)

# find maximum number of infections for each ensemble member
max_infections <- dat %>% 
                  group_by(fingerprint) %>%
                  summarize(max(I))

reorder_ix <- match(unique_fingerprint, max_infections$fingerprint)
max_infections <- max_infections[reorder_ix, ]

head(max_infections)
```


Plot each parameter against the output to get an idea of sensitivity

```{r sa_plot_univariate, fig.width=7, fig.height=7, out.width='70%'}
d <- ncol(X)
X.norm <- normalize(X)
y <- pull(max_infections,'max(I)')
X %>% 
  as_tibble %>% 
  mutate(y=y) %>% 
  gather('parameter', 'value', -y) %>% 
  ggplot(aes(x=value, y=y)) + 
    geom_point() + 
    facet_wrap(~parameter) +
    labs(y='output', x='input')
```

# Fit a Gaussian process emulator

```{r}
# Fit an emulator using DiceKriging
fit = km(~., design=X.norm, response=y)
```


## Leave-one-out cross validation of the fitted emulator

```{r sa_plot_loo, out.width='70%'}
loo = leaveOneOut.km(fit, type = 'UK', trend.reestim = TRUE)

tibble(y=y, em_mean=loo$mean, em_sd = loo$sd) %>%
  ggplot() + 
  geom_segment(aes(x=y, xend=y, y=em_mean - 2*em_sd, yend=em_mean + 2*em_sd)) +
  geom_point(aes(x=y, y=em_mean)) +
  geom_abline(intercept=-1, slope=1, lty=2) +
  labs(x='max. infections', y='emulator output')
```


## Perform a FAST99 sensitivity analysis 

cf. Saltelli et al (1999)

```{r}
# Generate a design for the FAST99 analysis
X.fast <- fast99(model = NULL, factors = colnames(X), n = 3000,
                 q = "qunif", q.arg = list(min = 0, max = 1))


# Predict the response at the FAST99 design points using the emulator
pred.fast = predict(fit, newdata = X.fast$X, type = 'UK')

# Calculate the sensitivity indices
fast.tell <- tell(X.fast, pred.fast$mean)

bp.convert <- function(fastmodel){
  # get the FAST summary into an easier format for barplot
  fast.summ <- print(fastmodel)
  fast.diff <- fast.summ[ ,2] - fast.summ[ ,1]
  fast.bp <- t(cbind(fast.summ[ ,1], fast.diff))
  fast.bp
}
```

```{r sa_plot_fast99, fig.width=7, fig.height=6, out.width='70%'}
par(las = 2, mar = c(9,5,3,2))
barplot(bp.convert(fast.tell), col = c('skyblue', 'grey'), 
        ylab = 'relative sensitivity', 
	main = 'FAST99 Sensitivity')
legend('topleft',legend = c('Main effect', 'Interactions'), 
       fill = c('skyblue', 'grey') )
```


# One-at-a-time sensitivity analysis

Parameters are swept across their range one at a time, with the remaining
parameters held at central values.

```{r}
n.oat <- 21
X.oat <- oaat.design(X.norm, n = n.oat, hold = rep(0.5,9))

colnames(X.oat) <- colnames(X)
pred.oat <- predict(fit, newdata = X.oat, type = 'UK')
```


```{r sa_plot_oat, fig.width=7, fig.height=8, out.width='70%'}
params = rep(colnames(X.oat), each=n.oat)
col_inds = rep(1:ncol(X.oat), each=n.oat)
tibble(parameter = params,
       value = X.oat[cbind(1:length(col_inds), col_inds)]) %>% 
  mutate(pred_mean=pred.oat$mean, 
	 pred_sd=pred.oat$sd,
	 lwr = pred_mean - 2 * pred_sd,
	 upr = pred_mean + 2 * pred_sd) %>% 
  ggplot(aes(x=value)) + 
    geom_ribbon(aes(ymin=lwr, ymax=upr), fill='gray') + 
    geom_line(aes(y=pred_mean)) + 
    facet_wrap(~parameter) +
    labs(x='Parameter value', y='Max. no. of Infections (emulator mean +/- 2 stdev)')
```


